Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-06-21 paper.tex
Fix bad mali figures

\subsubsection{Chained Scan, Decoupled Lookback}
Introduced by Merrill and Garland, the \emph{Decoupled Lookback} technique refines \emph{Chained Propagation} by separating reduction communication from dependency resolution, mitigating bottlenecks caused by propagation latency. In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction. Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback. However, granting each workgroup control over its dependency resolution does not eliminate reliance on \emph{progress guarantees}, as workgroups must still spin on tiles that have yet to post their reductions, risking starvation.

@@

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
\emph{Chained Scans} are \emph{one-way producer-consumer} algorithms; each workgroup consumes reductions only from \emph{lower-rank}\footnote{We specifically use the term \emph{rank} here to generalize between which ID---atomically acquired or scheduler-assigned---is used to determine this ordering. This distinction matters because the choice of ID determines which progress model, as defined by Sorensen et al.~\cite{sorensen2021}, is required to guarantee termination. Working under the assumption that the scheduler-assigned ID is used, Sorensen et al.~\cite{sorensen2021, sorensen2018} classify \emph{Chained Scan} as requiring \emph{HSA} (Heterogeneous System Architecture), a model that guarantees progress for the lowest-ranked workgroup. However, if the atomically acquired ID determines rank, then the HSA guarantee no longer suffices, as the lowest scheduler-assigned ID may not correspond to the lowest atomic ID. In this case, termination requires only \emph{OBE} (Occupancy-Bound Execution), which guarantees progress once a workgroup begins execution.} counterparts, thereby enforcing the ordering required to elide barriers. For such algorithms, guaranteeing execution of the lowest-rank workgroup is sufficient to ensure termination. If subgroups from the lowest-rank workgroup are not guaranteed execution, dependent workgroups may spin indefinitely, waiting for updates that never arrive. A \emph{forward-progress guarantee}, in the sense required by \emph{Chained Scan}, ensures that all occupant subgroups---potentially from different workgroups---are eventually granted execution after becoming active, thus preventing indefinite starvation.

@@

\begin{figure*}
  \centering
  \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
  \caption{\emph{Decoupled Fallback}. For brevity, the lookback is condensed; see Figure~\ref{fig:decoupled-lookback} for the full behavior, which is shared with \emph{Decoupled Lookback}.}
  \label{fig:decoupled-fallback}
\end{figure*}
\section{Decoupled Fallback}
\emph{Decoupled Fallback} addresses both inefficiencies. Our key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially. To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained Propagation} logic is typically performed by the first thread.

@@

\begin{table*}
  \centering
  \small
  \setlength{\tabcolsep}{1pt}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r r}
    \toprule
    Metric & \!\!\!\!\!\!Intel HD620 & ARM Mali-G78 MP20 & Apple M3 & Apple M1 Max & Nvidia 2080 Super & AMD 7900 XT \\
    \midrule
    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)      & 1.48    & 2.50     & 10.82   & 37.42    & 50.95    & 73.68    \\
    Avg \emph{DF} (32-bit ele/sec $\cdot 10^9$)          & 1.51    & 2.65     & 10.87   & 36.85    & 51.25    & 83.27    \\
    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)         & 1.06    & 1.94     & 7.46    & 25.75    & 34.36    & 62.57    \\
    \emph{DF}/\emph{Memcpy}                              & 102.0\% & 106.0\%  & 100.5\% & 98.4\%   & 100.6\%  & 113.0\%  \\
    \emph{DF}/\emph{RTS}                                 & 142.6\% & 136.6\%  & 145.7\% & 143.1\%  & 149.1\%  & 133.1\%  \\
    Avg Spins per workgroup                              & 0.420   & 2.52     & 0.719   & 0.752    & 0.679    & 0.533    \\
    Avg Lookback Length per workgroup                    & 1.714   & 3.52     & 4.598   & 10.737   & 12.350   & 22.648   \\
    Avg Total Fallbacks Initiated                        & 446.183 & 4739.91  & 449.961 & 1046.216 & 333.324  & 266.641  \\
    Avg Total Successful Insertions                      & 21.475  & 571.24   & 23.887  & 4.822    & 11.378   & 46.288   \\
    \bottomrule
  \end{tabular*}
  \caption{\emph{Decoupled Fallback} performance, Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements.\label{tab:results}}
\end{table*}

>>>>

\subsubsection{Chained Scan, Decoupled Lookback}
Introduced by Merrill and Garland~\cite{Merrill2016}, the \emph{Decoupled Lookback} technique refines \emph{Chained Propagation} by separating reduction communication from dependency resolution, mitigating bottlenecks caused by propagation latency. In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction. Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback. However, granting each workgroup control over its dependency resolution does not eliminate reliance on \emph{progress guarantees}, as workgroups must still spin on tiles that have yet to post their reductions, risking starvation.

@@

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
\emph{Chained Scans} are \emph{one-way producer-consumer} algorithms; each workgroup consumes reductions only from lower-\emph{rank} counterparts, thereby enforcing the ordering required to elide barriers. For such algorithms, guaranteeing execution of the lowest-rank workgroup is sufficient to ensure termination. If subgroups from the lowest-rank workgroup are not guaranteed execution, dependent workgroups may spin indefinitely, waiting for updates that never arrive. A \emph{forward-progress guarantee}, in the sense required by \emph{Chained Scan}, ensures that all occupant subgroups---from potentially different workgroups---are eventually granted execution after becoming active, thus preventing indefinite starvation.\footnote{We specifically use the term \emph{rank} here to generalize which ID---atomically acquired or scheduler-assigned---is used to determine this ordering. This distinction matters because the choice of ID determines which progress model, as defined by Sorensen et al.~\cite{sorensen2021}, is required to guarantee termination. Working under the assumption that the scheduler-assigned ID is used, Sorensen et al.~\cite{sorensen2021, sorensen2018} classify \emph{Chained Scan} as requiring \emph{HSA} (Heterogeneous System Architecture), a model that guarantees progress for the lowest-ranked workgroup. However, if the atomically acquired ID determines rank---as used by \emph{Decoupled Lookback}~\cite[Section 4.4]{Merrill2016}, \emph{StreamScan}~\cite[Section 3.3.4]{10.1145/2442516.2442539}, and this work---then the HSA guarantee no longer suffices, as the lowest scheduler-assigned ID may not correspond to the lowest atomic ID. In this case, termination requires \emph{OBE} (Occupancy-Bound Execution), which guarantees progress once a workgroup begins execution.}

@@

\begin{figure*}
  \centering
  \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
  \caption{\emph{Decoupled Fallback}. For brevity, the lookback is condensed; see Figure~\ref{fig:decoupled-lookback} for full lookback behavior.}
  \label{fig:decoupled-fallback}
\end{figure*}
\section{Decoupled Fallback}
\emph{Decoupled Fallback} addresses both inefficiencies. Our key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially. To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained Propagation} logic is typically performed by the first thread.

@@

\begin{table*}
  \centering
  \small
  \setlength{\tabcolsep}{1pt}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r r}
    \toprule
    Metric & \!\!\!\!\!\!Intel HD620 & ARM Mali-G78 MP20 & Apple M3 & Apple M1 Max & Nvidia 2080 Super & AMD 7900 XT \\
    \midrule
    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$) & 1.481   & 2.626   & 10.82   & 37.42    & 50.95   & 73.68   \\
    Avg \emph{DF} (32-bit ele/sec $\cdot 10^9$)     & 1.511   & 2.725   & 10.87   & 36.85    & 51.25   & 83.27   \\
    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)    & 1.06    & 2.025   & 7.46    & 25.75    & 34.36   & 62.57   \\
    \emph{DF}/\emph{Memcpy}                         & 102.0\% & 103.8\% & 100.5\% & 98.4\%   & 100.6\% & 113.0\% \\
    \emph{DF}/\emph{RTS}                            & 142.6\% & 134.6\% & 145.7\% & 143.1\%  & 149.1\% & 133.1\% \\
    Avg Spins per workgroup                         & 0.42    & 2.504   & 0.719   & 0.752    & 0.679   & 0.533   \\
    Avg Lookback Length per workgroup               & 1.714   & 3.519   & 4.598   & 10.737   & 12.35   & 22.648  \\
    Avg Total Fallbacks Initiated                   & 446.183 & 4702.01 & 449.961 & 1046.216 & 333.324 & 266.641 \\
    Avg Total Successful Insertions                 & 21.475  & 560.15  & 23.887  & 4.822    & 11.378  & 46.288  \\
    \bottomrule
  \end{tabular*}
  \caption{\emph{Decoupled Fallback} performance, Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements.\label{tab:results}}
\end{table*}
